### 2024-07-17 evaluating-prediction-accuracy.qmd
20240716 - reorder
20240717 - add more info

@@

To consider how we can evaluate the accuracy of predictions, consider an example adapted from @Meehl1955.
The military conducts a test of its prospective members to screen out applicants who would likely fail basic training.
To evaluate the accuracy of our predictions using the test, we can examine a confusion matrix.
A confusion matrix is a matrix that presents the predicted outcome on one dimension and the actual outcome (truth) on the other dimension.
If the predictions and outcomes are dichotomous, the confusion matrix is a 2x2 matrix with two rows and two columns that represent four possible predicted-actual combinations (decision outcomes): true positives (TP), true negatives (TN), false positives (FP), and false negatives (FN).

@@

Sensitivity is the proportion of those with the characteristic ($\text{TP} + \text{FN}$) that we identified with our measure ($\text{TP}$): $\frac{\text{TP}}{\text{TP} + \text{FN}} = \frac{86}{86 + 14} = .86$.
Specificity is the proportion of those who do not have the characteristic ($\text{TN} + \text{FP}$) that we correctly classify as not having the characteristic ($\text{TN}$): $\frac{\text{TN}}{\text{TN} + \text{FP}} = \frac{1,478}{1,478 + 422} = .78$.
PPV is the proportion of those who we classify as having the characteristic ($\text{TP} + \text{FP}$) who actually have the characteristic ($\text{TP}$): $\frac{\text{TP}}{\text{TP} + \text{FP}} = \frac{86}{86 + 422} = .17$
NPV is the proportion of those we classify as not having the characteristic ($\text{TN} + \text{FN}$) who actually do not have the characteristic ($\text{TN}$): $\frac{\text{TN}}{\text{TN} + \text{FN}} = \frac{1,478}{1,478 + 14} = .99$.

@@

There are four general patterns of miscalibration: overextremity, underextremity, overprediction, and underprediction (see @fig-miscalibration).
*Overextremity* exists when the predicted probabilites are too close to the extremes (zero or one).
*Underextremity* exists when the predicted probabilities are too far away from the extremes.
*Overprediction* exists when the predicted probabilities are consistently greater than the observed probabilities.
*Underprediction* exists when the predicted probabilities are consistently less than the observed probabilities.
For a more thorough description of these types of miscalibration, see @Lindhiem2020.\index{calibration}

>>>>

There are two primary dimensions of accuracy: (1) [discrimination](#sec-discrimination) and (2) [calibration](#sec-calibration).
Some general indexes of accuracy combine discrimination and calibration, as described in INSERT.

When dealing with a categorical outcome, discrimination is the ability to separate events from non-events.
When dealing with a continuous outcome, discrimination is the strength of the association between the predictor and the outcome.
Aspects of discrimination at a particular cutoff (e.g., sensitivity, specificity, area under the ROC curve) are described in INSERT.

There are four general patterns of miscalibration: overextremity, underextremity, overprediction, and underprediction (see @fig-miscalibration).
*Overextremity* exists when the predicted probabilites are too close to the extremes (zero or one).
*Underextremity* exists when the predicted probabilities are too far away from the extremes.
*Overprediction* exists when the predicted probabilities are consistently greater than the observed probabilities.
*Underprediction* exists when the predicted probabilities are consistently less than the observed probabilities.
For a more thorough description of these types of miscalibration, see @Lindhiem2020.

@@

To consider how we can evaluate the accuracy of predictions for a categorical outcome, consider an example adapted from @Meehl1955.
The military conducts a test of its prospective members to screen out applicants who would likely fail basic training.
To evaluate the accuracy of our predictions using the test, we can examine a confusion matrix.
A confusion matrix is a matrix that presents the predicted outcome on one dimension and the actual outcome (truth) on the other dimension.
If the predictions and outcomes are dichotomous, the confusion matrix is a 2x2 matrix with two rows and two columns that represent four possible predicted-actual combinations (decision outcomes): true positives (TP), true negatives (TN), false positives (FP), and false negatives (FN).

@@

Sensitivity is the proportion of those with the characteristic ($\text{TP} + \text{FN}$) that we identified with our measure ($\text{TP}$): $\frac{\text{TP}}{\text{TP} + \text{FN}} = \frac{86}{86 + 14} = .86$.
Specificity is the proportion of those who do not have the characteristic ($\text{TN} + \text{FP}$) that we correctly classify as not having the characteristic ($\text{TN}$): $\frac{\text{TN}}{\text{TN} + \text{FP}} = \frac{1,478}{1,478 + 422} = .78$.
PPV is the proportion of those who we classify as having the characteristic ($\text{TP} + \text{FP}$) who actually have the characteristic ($\text{TP}$): $\frac{\text{TP}}{\text{TP} + \text{FP}} = \frac{86}{86 + 422} = .17$.
NPV is the proportion of those we classify as not having the characteristic ($\text{TN} + \text{FN}$) who actually do not have the characteristic ($\text{TN}$): $\frac{\text{TN}}{\text{TN} + \text{FN}} = \frac{1,478}{1,478 + 14} = .99$.

@@
